noam_beta_diversity

### 2024-09-04 noam_beta_diversity.Rmd
finished nmds on macroalgae, did permanova, modeled beta dispersion

@@

# transformed dissimilarities
bray_time_series_site_macro$dissimilarity_trans <- sv_trans(bray_time_series_site_macro$dissimilarity, bray_time_series_site_macro)

@@

@@

full_site_species_matrix <- full_site_species_matrix[rownames(full_site_species_matrix) %in% relevant_sites,] # subset so only quads that didn't contain rare species are represented
full_site_species_matrix <- asin(sqrt(full_site_species_matrix)) # arcsin squareroot transform
site_full$stratum <- paste(site_full$year,"-", site_full$location, sep = "") # for merging with bray matrix later
full_subset <- site_full[site_full$stratum %in% relevant_sites,] # make sure rows are also removed from dataframe

@@

full_site_macro_species_matrix <- as.matrix(site_macro[ , -which(names(site_macro) %in% meta_cols)])
full_site_macro_species_matrix <- full_site_macro_species_matrix/site_macro$hard_cover # convert from percent to proportion of hard cover
rownames(full_site_macro_species_matrix) <- paste(site_macro$year,"-", site_macro$site_habitat, sep = "")

full_site_macro_species_matrix <- full_site_macro_species_matrix[rownames(full_site_macro_species_matrix) %in% relevant_site_macros,] # subset so only quads that didn't contain rare species are represented
full_site_macro_species_matrix <- asin(sqrt(full_site_macro_species_matrix)) # arcsin squareroot transform
site_macro$stratum <- paste(site_macro$year,"-", site_macro$location, sep = "") # for merging with bray matrix later
full_subset_site_macro <- site_macro[site_macro$stratum %in% relevant_site_macros,] # make sure rows are also removed from dataframe

@@

# read in disturbances
disturbances <- read.csv(here("data", "disturbance_list.csv"))
full.cent_site_macro <- merge(full.cent_site_macro, disturbances, by = "year") # add disturbances to path

ggplot() + 
  geom_point(data= full.site.scores_site_macro, aes(x=NMDS1,y=NMDS2, colour = habitat), alpha = 0.05) + # add the point markers
  geom_text_repel(data= full.abundant.species.scores_site_macro,aes(x=NMDS1,y=NMDS2, label = species), size = 4, alpha = 0.5) +  # add the species labels

  geom_path(data = full.cent_site_macro, aes(x = NMDS1, y = NMDS2, group = habitat, colour = habitat),
            arrow = arrow(length=unit(0.25,"cm"), ends="last", type = "closed"), size = 1) +
  geom_point(data = full.cent_site_macro, aes(x = NMDS1, y = NMDS2, group = habitat, colour = habitat, shape = disturbance), size = 3, stroke = 1.4) +
  scale_shape_manual(values = c(
    "cots" = 8,
    "cyclone" = 13,
    "bleaching" = 4
  )) +

@@

>>>>

```{r dissimilarity compared to baseline all, fig.height=15, fig.width = 15}
bray_time_series_site_baseline <- subset(bray_time_series_site, year2 == 2007)
bray_mod_site_baseline <- glmmTMB(dissimilarity_trans ~ time*habitat + (1 | site),
                              na.action = na.pass, family = beta_family(link = "logit"), data = bray_time_series_site_baseline)
hist(residuals(bray_mod_site_baseline)) # long tails but not bad
plot(residuals(bray_mod_site_baseline) ~ fitted(bray_mod_site_baseline) ) # heteroskedastic

emtrends(bray_mod_site_baseline, pairwise ~ habitat , "time") # fringing slope significantly different from all others
bray_trends_emm_site_baseline <- emmip(bray_mod_site_baseline, habitat ~ time , at = list(time = seq(1,16,1)), plotit = F, CIs = T)

ggplot()+
   geom_point(data = bray_time_series_site_baseline, aes(x = time, y = dissimilarity, colour = habitat), alpha = 0.1 ) +
   geom_line(data = bray_trends_emm_site_baseline, aes(x = time, y = inv_logit(yvar), colour = habitat), linewidth = 2) +
   geom_ribbon(data = bray_trends_emm_site_baseline, aes(x = time, ymin = inv_logit(LCL), ymax = inv_logit(UCL), fill = habitat), alpha = 0.3) +
   scale_colour_manual(values = habitat_colours) +
   scale_fill_manual(values = habitat_colours) +

@@

# model
# transformed dissimilarities
bray_time_series_site_macro$dissimilarity_trans <- sv_trans(bray_time_series_site_macro$dissimilarity, bray_time_series_site_macro)

@@

```{r dissimilarity compared to baseline macro, fig.height=15, fig.width = 15}
bray_time_series_site_macro_baseline <- subset(bray_time_series_site_macro, year2 == 2007)
bray_mod_site_macro_baseline <- glmmTMB(dissimilarity_trans ~ time*habitat + (1 | site),
                              na.action = na.pass, family = beta_family(link = "logit"), data = bray_time_series_site_macro_baseline)
hist(residuals(bray_mod_site_macro_baseline)) # slight negative skew
plot(residuals(bray_mod_site_macro_baseline) ~ fitted(bray_mod_site_macro_baseline) ) # strong negative trend

emtrends(bray_mod_site_macro_baseline, pairwise ~ habitat , "time") # fringing slope significantly different from all others
bray_trends_emm_site_macro_baseline <- emmip(bray_mod_site_macro_baseline, habitat ~ time , at = list(time = seq(1,16,1)), plotit = F, CIs = T)

ggplot()+
   geom_point(data = bray_time_series_site_macro_baseline, aes(x = time, y = dissimilarity, colour = habitat), alpha = 0.1 ) +
   geom_line(data = bray_trends_emm_site_macro_baseline, aes(x = time, y = inv_logit(yvar), colour = habitat), linewidth = 2) +
   geom_ribbon(data = bray_trends_emm_site_macro_baseline, aes(x = time, ymin = inv_logit(LCL), ymax = inv_logit(UCL), fill = habitat), alpha = 0.3) +
   scale_colour_manual(values = habitat_colours) +
   scale_fill_manual(values = habitat_colours) +

@@

full_site_species_matrix <- full_site_species_matrix[rownames(full_site_species_matrix) %in% relevant_sites,] # subset so only quads that didn't contain rare species are represented
full_site_species_matrix <- asin(sqrt(full_site_species_matrix)) # arcsin squareroot transform
site_full$stratum <- paste(site_full$year,"-", site_full$site_habitat, sep = "") # for merging with bray matrix later
full_subset <- site_full[site_full$stratum %in% relevant_sites,] # make sure rows are also removed from dataframe

@@

full_site_macro_species_matrix <- as.matrix(site_macro[ , -which(names(site_macro) %in% colnames(site_meta_df))])
full_site_macro_species_matrix <- full_site_macro_species_matrix/site_macro$hard_cover # convert from percent to proportion of hard cover
rownames(full_site_macro_species_matrix) <- paste(site_macro$year,"-", site_macro$site_habitat, sep = "")

full_site_macro_species_matrix <- full_site_macro_species_matrix[rownames(full_site_macro_species_matrix) %in% relevant_site_macros,] # subset so only quads that didn't contain rare species are represented
full_site_macro_species_matrix <- asin(sqrt(full_site_macro_species_matrix)) # arcsin squareroot transform
site_macro$stratum <- paste(site_macro$year,"-", site_macro$site_habitat, sep = "") # for merging with bray matrix later
full_subset_site_macro <- site_macro[site_macro$stratum %in% relevant_site_macros,] # make sure rows are also removed from dataframe

@@

# read in disturbances
full.cent_site_macro <- merge(full.cent_site_macro, disturbances, by = "year") # add disturbances to path

ggplot() + 
  geom_point(data= full.site.scores_site_macro, aes(x=NMDS1,y=NMDS2, colour = habitat), alpha = 0.2) + # add the point markers
  geom_text_repel(data= full.abundant.species.scores_site_macro,aes(x=NMDS1,y=NMDS2, label = species), size = 6, alpha = 0.7) +  # add the species labels

  geom_path(data = full.cent_site_macro, aes(x = NMDS1, y = NMDS2, group = habitat, colour = habitat),
            arrow = arrow(length=unit(0.4,"cm"), ends="last", type = "closed"), size = 1) +
  geom_point(data = full.cent_site_macro, aes(x = NMDS1, y = NMDS2, group = habitat, colour = habitat, shape = disturbance), size = 5, stroke = 1.4) +
  scale_shape_manual(values = c(
    "cots" = 8,
    "cyclone" = 13,
    "bleaching" = 4
  )) +

@@

```{r full site permanova}
full_site_bray <- vegdist(full_site_species_matrix, "bray") 
full_site_macro_bray <- vegdist(full_site_macro_species_matrix, "bray")

# adonis2(full_site_bray ~ year*habitat, data = site_full)
#               Df SumOfSqs      R2       F Pr(>F)    
# year           1    1.293 0.03257 18.4422  0.001 ***
# habitat        3    9.952 0.25065 47.3113  0.001 ***
# year:habitat   3    1.255 0.03161  5.9659  0.001 ***
# Residual     388   27.204 0.68518                   
# Total        395   39.704 1.00000                   
# habitat accounts for most of the variation in algal communities

# adonis2(full_site_macro_bray ~ year*habitat, data = full_subset_site_macro)
#               Df SumOfSqs      R2       F Pr(>F)    
# year           1    6.209 0.04528 24.8046  0.001 ***
# habitat        3   28.410 0.20719 37.8323  0.001 ***
# year:habitat   3    5.882 0.04290  7.8332  0.001 ***
# Residual     386   96.622 0.70464                   
# Total        393  137.124 1.00000               
# same when just looking at macroalgal communities, but still greater residual variance

```{r beta dispersion all algae, fig.height=15, fig.width=15}
full_site_betadisper <- vegan::betadisper(full_site_bray, full.site.scores_site$hab_year)
site_full$betadispersion <- full_site_betadisper$distances[match(site_full$stratum, names(full_site_betadisper$distances))]
site_full$betadispersion_trans <- sv_trans(site_full$betadispersion, site_full)

betadisper_site_mod <- glmmTMB(betadispersion_trans ~ year*habitat + (1|site), family = beta_family(link = "logit"), data = site_full)
summary(betadisper_site_mod) # fair bit of within site variance - inv_logit(0.04) = 0.5
hist(residuals(betadisper_site_mod)) # pretty good
plot(residuals(betadisper_site_mod) ~ fitted(betadisper_site_mod)) # slight heteroskedasticity
Anova(betadisper_site_mod)
#                Chisq Df Pr(>Chisq)    
# year         12.9561  1  0.0003189 ***
# habitat      27.2191  3  5.296e-06 ***
# year:habitat  6.8472  3  0.0769311 .

emtrends(betadisper_site_mod, pairwise ~ habitat , "year") # fringing slope significantly different from outer17
betadisper_site_emm <- emmip(betadisper_site_mod, habitat ~ year , at = list(year = seq(2007,2023,1)), plotit = F, CIs = T)

ggplot()+
   geom_point(data = site_full, aes(x = year, y = betadispersion_trans, colour = habitat), alpha = 0.1 ) +
   geom_line(data = betadisper_site_emm, aes(x = year, y = inv_logit(yvar), colour = habitat), linewidth = 2) +
   geom_ribbon(data = betadisper_site_emm, aes(x = year, ymin = inv_logit(LCL), ymax = inv_logit(UCL), fill = habitat), alpha = 0.3) +
   scale_colour_manual(values = habitat_colours) +
   scale_fill_manual(values = habitat_colours) +

   labs(y = "Beta dispersion\n", x = "Year") +
   theme_bw() + 
   theme(panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(),
         panel.background = element_blank(),
         plot.title = element_text(color = "grey20", size = 32, vjust = 0, face = "plain"),
         axis.line.y = element_line(colour = "black"),
         axis.text.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = .5, face = "plain"),
         axis.title.y = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
         axis.line.x = element_line(colour = "black"),
         axis.text.x = element_text(color = "grey20", size = 28, angle = 90, vjust = 0.5, hjust=1, face = "plain"),
         axis.title.x = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
         strip.text.x = element_text(size = 32),
         legend.key.size = unit(2, "cm"), # increase legend size
         legend.text = element_text(size = rel(2)), # increase legend text
         legend.title = element_text(size = rel(2.25)),
         legend.position = 'bottom', # move legend to bottom
         legend.box = 'horizontal',
         legend.box.background = element_rect(colour = "black") )# make legend horizontal

```{r beta dispersion macoralgae, fig.height=15, fig.width=15}
full_site_macro_betadisper <- vegan::betadisper(full_site_macro_bray, full.site.scores_site_macro$hab_year)
full_subset_site_macro$betadispersion <- full_site_macro_betadisper$distances[match(full_subset_site_macro$stratum, names(full_site_macro_betadisper$distances))]
full_subset_site_macro$betadispersion_trans <- sv_trans(full_subset_site_macro$betadispersion, full_subset_site_macro)
